---
output:
  pdf_document: default
  html_document: default
---
# Football Events Project
### Summer 2021

**Required Packages:**

- tidyverse
- ggplot2

---

To find the data used in this file, see "GitHub Preparing Data.R" in R Code folder.
  
The rest of this analysis will use the events_new.csv file which was updated in the aforementioned data prep code.

```{r, message=F, warning=F, echo=F}
library(tidyverse)
events = read.csv("C:/Users/schol/OneDrive/Documents/Summer 21 Project/events_new.csv") # Edit file location as necessary

events = events %>% mutate(
  event_type = as.factor(event_type),
  event_type2 = as.factor(event_type2),
  side = as.factor(side),
  shot_place = as.factor(shot_place),
  shot_outcome = as.factor(shot_outcome),
  is_goal = as.factor(is_goal),
  location = as.factor(location),
  bodypart = as.factor(bodypart),
  assist_method = as.factor(assist_method),
  situation = as.factor(situation),
  fast_break = as.factor(fast_break)
)
```

---

```{r}
summary(events)
```
  
The dataset contains 24 variables and 941,009 observations, where each observation is an event taking place in a match. To perform some initial exploratory analysis into the variables individually, I will go through those for which this makes the most sense and display information about them through graphics.

---

### Time:
  
The time variable gives the minute in which the given event took place and ranges from 0-100.

```{r, message=F, warning=F}
library(ggplot2)
ggplot(events) + geom_bar(aes(x = time), fill = "dark blue", width = 1) + 
  labs(x = "Minute", y = "Number of Events", title = "Number of Events in each Minute") +
  annotate("text", x = 45, y = 25000, label = "45'") +
  annotate("text", x = 90, y = 44000, label = "90'")
```

There appears to be a slight upward trend from 0-90 with respect to the number of events occurring in each minute. Moreover, there are two spikes in the plot which correspond to the 45th and 90th minute. This is likely because some events which happened during injury time, e.g. the 45+2 minute, may have been coded as the 45th minute, and the same for the 90th minute. The fact that the spike in the 90th minute is larger than that for the 45th minute is probably because there tends to be more time added on at the end of the second half of a game than the first, mostly because of substitutions or injuries when players get tired.

---

### Event type:
  
event_type shows what occurred in each event, with 12 possible event types.

```{r}
event_type.labs = c("0" = "Announcement", "1" = "Attempt", "2" = "Corner", "3" = "Foul", "4" = "Yellow card",
                    "5" = "Second yellow card", "6" = "Red card", "7" = "Substitution", "8" = "Free kick won", 
                    "9" = "Offside", "10" = "Hand ball", "11" = "Penalty conceded")

events %>% add_count(event_type) %>% ggplot() +
  geom_bar(aes(y = reorder(event_type, n)), fill = "dark red") + 
  scale_y_discrete("Event Type", labels = event_type.labs) +
  labs(x = "Number of Events", title = "Number of Events by Type")
```

The three most common events by a large margin are free kicks won, fouls, and attempts respectively, each with over 200,000 events. The next most common event is corners, with almost 100,000 events. The rest of the event types are around 50,000 occurrences or less, with the lowest event types being penalties conceded and sending offs (red card and second yellow card), with there being no second yellow cards in the data. Generally, this suggests that the best events to analyse would be fouls (including free kicks won) or attempts, since these are the events for which there is the most data.
  
  
There is a second variable for event types, which contains additional information about attempts, yellow cards, second yellow cards, red cards, and offsides.

```{r}
event_type2.labs = c("12" = "Key pass", "13" = "Failed through ball", "14" = "Sending off", "15" = "Own goal")

event_type2.labs = c("12" = "Key pass", "13" = "Failed through ball", "14" = "Sending off", "15" = "Own goal")
events[!(is.na(events$event_type2)),] %>% add_count(event_type2) %>% ggplot() +
  geom_bar(aes(y = reorder(event_type2, n)), fill = "dark blue") +
  scale_y_discrete("Event Type", labels = event_type2.labs) +
  labs(x = "Number of Events", title = "Number of Events by Type 2")
```

Most of the observations for event_type2 are NA, because they are characterised by one of the events in the event_type category, therefore NAs have been excluded in this plot. The most common event in this category is key passes by a large margin, followed by a failed through ball.

---

### Event team:

Given the large number of teams in the dataset, I will take a subset of the most common teams and look at which teams have the most events.

```{r}
events %>% add_count(event_team) %>% filter(n > 10000) %>% ggplot() +
  geom_bar(aes(y = reorder(event_team, n), fill = league)) +
  scale_fill_discrete("League") +
  labs(y = "Team", x = "Number of Events", title = "Number of Events per Team for Common Teams")
```

The plot shows the number of events per team for teams with more than 10,000 events in the data. Interestingly, most of the top 10 teams are from the Serie A (Italy), even though they play the same number of games per season as Ligue 1 and La Liga. The Bundesliga teams only occur at the lower end of the plot, which is due to there being fewer games per season in the Bundesliga.
  
```{r}
events %>% add_count(event_team) %>% filter(n > 10000) %>% ggplot() +
  geom_bar(aes(y = reorder(event_team, n), fill = event_type)) +
  scale_fill_discrete("Event Type", labels = event_type.labs) +
  labs(y = "Team", x = "Number of Events", title = "Number of Events per Team for Common Teams")
```
  
The breakdown of events for common teams by type shows what was shown in the previous plot of event types: that the most common types are attempts, fouls, and free kicks won. Thereafter, there is some fluctuation with some teams having more corners and others having more offsides, as is expected with different teams using different strategies and tactics.


---

### Player:

Just as there are too many teams to see how many events they have, there are even more players in the data, so a similar approach will be taken as above and the most common players by number of events will be considered.

```{r}
events[!(is.na(events$player)) & !(is.na(events$event_type)),] %>% add_count(player) %>% filter (n > 1000) %>%
  ggplot() + geom_bar(aes(y = reorder(player, n), fill = event_type)) +
  scale_fill_discrete("Event Type", labels = event_type.labs) +
  labs(y = "Player", x = "Number of Events", title = "Number of Events per Player for Common Players")
```

Common players are filtered as those who have more than 1,000 events. The most frequently occuring players are typically world class players who are forwards and therefore have a lot of attempts. However, there are some notable exceptions in the plot. Stefan Kiessling, a centre forward who played in the Bundesliga, is the fifth most frequent player in the data, but has far fewer attempts than the players above him. Kiessling's events were mostly winning free kicks and fouling other players, and also having attempts. Juanfran is another interesting exception, given he is a defender, but he clearly took the corners for his team for a long time with more corners than attempts.

---

### Shot place and bodypart:

Given that shots will always be with a bodypart, it makes sense to look at these two variable together and see which shot placements are most common and what the breakdown is by bodypart.

```{r}
shot_place.labs = c("1" = "Bit too high", "2" = "Blocked", "3" = "Bottom left corner", "4" = "Bottom right corner",
                    "5" = "Centre of the goal", "6" = "High and wide", "7" = "Hits the bar", 
                    "8" = "Misses to the left", "9" = "Misses to the right", "10" = "Too high", 
                    "11" = "Top centre of the goal", "12" = "Top left corner", "13" = "Top right corner")
bodypart.labs = c("1" = "Right foot", "2" = "Left foot", "3" = "Head")

events[!(is.na(events$shot_place)),] %>% add_count(shot_place) %>% ggplot() +
  geom_bar(aes(y = reorder(shot_place, n), fill = bodypart)) +
  scale_y_discrete("Shot Place", labels = shot_place.labs) +
  scale_fill_discrete("Bodypart", labels = bodypart.labs) +
  labs(x = "Number of Shots", title = "Shot placement and Bodypart Used")
```

By far the most common shot placement is that the shot is blocked, so the placement cannot be identified. Thereafter, shots usually miss to the left or right or go in the centre of the goal, then the bottom corners, and then shots which are too high and wide. For all of the shot placements, the most common bodypart is the right foot, typically followed by left foot and then header. This makes sense since it is most common for players to be right footed, and headers will not happen as common as shots with feet.
  
---

### Location:

Location refers to the area on the pitch in which an event took place. The labels of Location can be somewhat unclear because some can include others, for example it appears that most of the locations could be classed as in the attacking half, which is itself a location. Looking at the breakdown of event types by location explains why this is.

```{r}
location.labs = c("1" = "Attacking half", "2" = "Defensive half", "3" = "Centre of the box", "4" = "Left wing",
                  "5" = "Right wing", "6" = "Difficult angle and long range", "7" = "Difficult angle on the left",
                  "8" = "Difficult angle on the right", "9" = "Left side of the box", "10" = 
                    "Left side of the six yard box", "11" = "Right side of the box", "12" = 
                    "Right side of the six yard box", "13" = "Very close range", "14" = "Penalty spot",
                  "15" = "Outside of the box", "16" = "Long range", "17" = "More than 35 yards", "18" = 
                    "More than 40 yards", "19" = "Not recorded")

events[!(is.na(events$location)),] %>% add_count(location) %>% ggplot() +
  geom_bar(aes(y = reorder(location, n), fill = event_type)) +
  scale_y_discrete("Location", labels = location.labs) + 
  scale_fill_discrete("Event Type", labels = event_type.labs) +
  labs(x = "Number of Events", title = "Number of Events by Location on Pitch")
```

Looking at the breakdown of locations by event type, we see that only two event types have a location attached to them, which are attempts and free kicks won. Moreover, some of the location tags correspond to attempts and some to free kicks won but none to both. For free kicks won, the most common location is defensive half, followed by attacking half, and then left and right wing. It is unclear whether events on the left and right wing are in the attacking and defensive half, or whether any events in either half are on the wings given a lack of information about the variables. However, the combination of attacking half, left wing, and right wing would be close to the number of events in the defensive half, so the wings could be considered in the attacking half. For attempts, most of them occur outside of the box and then from the centre of the box. Then, there is a large gap before right and left side of the box before the rest, which have a small number of events relatively speaking.
  
In terms of attempts, most of them come from outside of the box followed by the centre of the box, with the rest being comparatively infrequent. It is interesting to look at how many attempts resulted in goals, shown below.

```{r}
is_goal.labs = c("0" = "No Goal", "1" = "Goal")
events[events$event_type == 1,] %>% add_count(location, sort = T) %>%
  ggplot() + geom_bar(aes(y = reorder(location, n), fill = is_goal)) +
  scale_y_discrete("Location", labels = location.labs) +
  scale_fill_discrete("Outcome", labels = is_goal.labs) +
  labs(x = "Number of Events", title = "Proportion of Attempts Resulting in Goals by Location")
```

While there were quite a few more attempts from outside of the box, far fewer resulted in a goal when compared to those from the centre of the box, as is to be expected.

---

### Assist method:

This variable gives, for every goal, what the type of assist was that came before the goal, meaning the last action by a teammate before the goal.

```{r}
assist_method.labs = c("0" = "None", "1" = "Pass", "2" = "Cross", "3" = "Headed pass", "4" = "Through ball")

events[!(is.na(events$assist_method)),] %>% add_count(assist_method) %>% ggplot() +
  geom_bar(aes(y = reorder(assist_method, n), fill = is_goal)) +
  scale_y_discrete("Assist Method", labels = assist_method.labs) +
  scale_fill_discrete("Goal Scored", labels = c("No", "Yes")) +
  labs(x = "Number of Events", title = "Assist Methods")
```

The overwhelming majority of goals are categorized as without an assist, meaning the last player to touch the ball before the attempt was an opposition player or the attacker intercepted / tackled to gain possession before the attempt without passing. Moreover, it could be that some attempts which had assists were not coded and were simply assigned to none as opposed to the correct assist. There are some ambiguities with this variable caused by a lack of explanation for distinction between levels. For example, it is unclear what constitutes a pass compared to a cross or through ball. This ambiguity makes it difficult to perform analysis with the variable because interpretation will be difficult. Filling the bars by number of goals scored, we see that assists apply even when attempts aren't converted, which is the case in the vast majority of attempts.

